Shaping strategies and standardisation of faceted spheroids from Qianshangying (North China)

### 2025-06-10 analysis/paper/paper.qmd
editing results section and finessing a few tables and plots

@@

## Core Type and Typology

@@

The diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising the following phases:

Phase 1) unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation. Large and flat surfaces of the cobble are selected as platforms to striking toward the opposite surface. Knapping proceeds until the peripheral volume is fully exploited, resulting in an approximately cubic form. Phase 2) peripheral centripetal exploitation. In this phase, previous removals in phase 1 serve as platforms for centripetal flaking around the core periphery. This represents a surface exploitation strategy. A notable feature is the angle between the face ventral surface and platform (core edge angle), which often exceeds 90°, forming what the literature refers to as secant plan rather than parallel plans. This approach contributes significantly to the progressive spherical of the shape. Phase 3) small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericality. It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the initial visible sequence are often obliterated. We hypothesize the existence of a Phase 0, potentially involving roughing-out of the cobble or an earlier cycle of Phases 1–3.

Scar orientation analysis further supports the interpretation of a systematical reduction sequence in the production of QSY spheroids. Shown in the ternary plot, spheroids tend to cluster together, with elongation ratios ranging from 0.403 to 0.637 and isotropy ratios from 0.164 to 0.497. These elongation ratios suggest that the spheroids underwent unidirectional or bidirectional total exploitation, resulting in relatively high elongation values. In addition, the peripheral centripetal exploitation appears to have contributed to a certain degree of isotropy in scar arrangement—higher isotropy ratios are associated with more spherical shapes. Subspheroids exhibit greater variability in both elongation and isotropy ratios, indicating less consistent reduction intensity.

@@

# 检查结果
print(head(benn.output))

@@

```{r, Fig, benn_ternary_plot, echo=FALSE}

@@

Based on this diacritical analysis and scar orientation analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY. This strategy consists of a series of essential and sequential phases: a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases. The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids—although some are relatively large—mostly cluster around a diameter of 8–10 cm. This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 8–10 cm was considered an optimal or intended size in the knapping process.

## Efficiency of shaping spheroids

@@

file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
data_raw2 <- read_excel(file_path)
knitr::kable(data_raw2, caption = "Summary of core typology and morphological parameters.")

@@

```{r, Fig spearman correlation matrix, echo=FALSE}

@@

# 显示维度和列名确认
print(dim(selected_cols))
print(colnames(selected_cols))

@@

# 打印结果查看
print("Spearman correlation matrix:")
print(round(cor_matrix, 4))
print("P-value matrix:")
print(round(p_matrix, 4))

@@

# 显著标记星号
labels_long$star <- ifelse(labels_long$p < 0.1 & labels_long$Var1 != labels_long$Var2, "*", "")

@@

# 自己绘图
ggplot(labels_long, aes(x = Var2, y = Var1, fill = cor)) +
  geom_tile(color = "gray") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  geom_text(aes(label = cor_label), size = 3) +          
# 相关系数字符串
  geom_text(aes(label = star), color = "red", size = 6, nudge_y = 0.25) + # 星号
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Spearman Correlation Matrix with Significance",
       fill = "Correlation")

@@

The Spearman correlation analysis shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI. Among them, scar counts show significant positive correlations with both removal ratio (ρ = `r cor_scar_removal`, p = $`r p_scar_removal`$) and SDI (ρ = `r cor_scar_sdi`, p = $`r p_scar_sdi`$). Removal ratio and SDI exhibit a stronger and highly significant correlation (ρ = `r cor_removal_sdi`, p = $`r p_removal_sdi`$). These two variables incorporate surface area in their calculation, making them more effective in capturing the density and intensity of reduction, and therefore more reliable indicators of the shaping process in spheroids.

Among the shape descriptors of spheroids, several significant correlations were observed. Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with SPHARM power (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms. As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids. This may help explain why spheroids were often interpreted as exhausted cores in earlier research. Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult. However, the reverse is not necessarily true—not all exhausted cores evolve into spheroids.

Curvature was only significantly correlated with sphericity (ρ = `r cor_cur_sph`, p = $`r p_cur_sph`$), suggesting that smoother surfaces tend to approximate spheres more closely. Its lack of correlation with other variables may reflect its focus on local surface texture, while edge angle and SPHARM power capture broader aspects of shape regularity and symmetry.

The weak correlation between sphericity and SPHARM power may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SPHARM power responds more to asymmetry and irregularity. Thus, even in similarly spherical specimens, surface differences can produce substantial variation in SPHARM power.

The analysis of spheroid shape variables indicates that mean edge angle, curvature, sphericity, and SPHARM power each capture distinct aspects of spheroid morphology, which explains the varying strengths of correlation observed among them.

<!-- Here's some example analysis code: -->

```{r}
#| label: fig-m
#| fig-cap: "the"
#| out-width: "100%"

knitr::include_graphics(here("analysis/paper/figures/1.png"))
```

@fig-m shows...

>>>>

... Core Type and Typology ...

<!--  introduce some key concepts that we use later such as SSDA -->

@@

@@

The diacritical analysis of spheroid surfaces reveals a structured and recurrent reduction sequence comprising the following phases. Phase 1 was a unidirectional or bidirectional total exploitation, belonging to one kind of volume exploitation. Large and flat surfaces of the cobble were selected as platforms for striking toward the opposite surface. Knapping proceeded until the peripheral volume was fully exploited, resulting in an approximately cubic form. Phase 2 was peripheral centripetal exploitation. In this phase, previous removals in phase 1 serve as platforms for centripetal flaking around the core periphery. This represents a surface exploitation strategy. A notable feature is the angle between the face's ventral surface and platform (i.e. core edge angle), which often exceeds 90°, forming a secant plan [citation] rather than parallel planes. This phase contributes significantly to the increasingly spherical shape of the artefact. Phase 3 was the production of small scars on ridges or intersections of previous flake scars, resulting in a rounding and smoother surface, further enhancing sphericality. It is important to note that due to the high degree of reduction intensity observed in most spheroids, phases prior to the visible sequence were often obliterated. We hypothesize the existence of a Phase 0, potentially involving roughing-out of the cobble or an earlier cycle of Phases 1–3.

<!-- insert QSYA-1098-4.png and QSYA-2797-0.png about here, with the texure paired with the phase annotations figures, like a 2x2 panel, add a legend to indicate the meaning of the colours -->

Scar orientation analysis further supports the interpretation of a systematic reduction sequence in the production of QSY spheroids. @fig-benn-ternary-plot shows that spheroids tend to cluster together, with elongation ratios ranging from 0.403 to 0.637 and isotropy ratios from 0.164 to 0.497. These elongation ratios suggest that the spheroids underwent unidirectional or bidirectional total exploitation, typical of Phase 1, resulting in relatively high elongation values. In addition, the peripheral centripetal exploitation, namely Phase 2, appears to have contributed to a certain degree of isotropy in scar arrangement—higher isotropy ratios are associated with more spherical shapes. Subspheroids exhibit greater variability in both elongation and isotropy ratios, indicating less consistent reduction intensity.

@@

# 检查结果
# print(head(benn.output))

@@

```{r, echo=FALSE}
#| label: fig-benn-ternary-plot
#| fig-cap: "Ternary plot showing flake scar orientation analysis"

@@

<!-- Based on this diacritical analysis and scar orientation analysis, we can summarize the technical strategy employed in the production of faceted spheroids in QSY. This strategy consists of a series of sequential phases: a potential rough-out phase, volume exploitation aimed at forming a cubic shape, surface exploitation to approximate a spheroidal geometry, edge refinement, and potentially the repetition of earlier phases. The existence of a repetitive phase is supported by the observation that while the original cobble blanks in QSY display considerable size variation, the finished spheroids—although some are relatively large—mostly cluster around a diameter of 8–10 cm. This suggests that the production of spheroids may have involved multiple cycles of reduction, and further implies that a diameter of 8–10 cm was considered an optimal or intended size in the knapping process. BM: maybe move to discussion -->

<!-- consider to add a visualisation, such as a histogram of the diameter, to support this claim, maybe above we can have a panel of plots of several basic attributes just to introduce the assemblages to the reader, then refer to it here. Maybe also some representative photos-->

@@

file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
data_raw2 <- 
  read_excel(file_path) %>% 
  drop_na() %>% 
  mutate(across(where(is.numeric), \(x) round(x, 3)))

knitr::kable(data_raw2, caption = "Summary of core typology and morphological parameters.")

@@

```{r, echo=FALSE}
#| label: fig-spearman-correlation-matrix
#| fig-cap: "Spearman correlation matrix of shape variables. Red star indicates p < 0.05"

@@

# 显示维度和列名确认
# print(dim(selected_cols))
# print(colnames(selected_cols))

@@

# 打印结果查看
# print("Spearman correlation matrix:")
# print(round(cor_matrix, 4))
# print("P-value matrix:")
# print(round(p_matrix, 4))

@@

# 显著标记星号
labels_long$star <- ifelse(labels_long$p < 0.05 & labels_long$Var1 != labels_long$Var2, "*", "")

@@

# 自己绘图
ggplot(labels_long, aes(x = Var2, y = Var1, fill = cor)) +
  geom_tile(color = "gray") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  geom_text(aes(label = cor_label), size = 3) +          
# 相关系数字符串
  geom_text(aes(label = star), color = "red", size = 6, nudge_y = 0.25) + # 星号
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(fill = "Correlation",
       x = "",
       y = "")

@@

@fig-spearman-correlation-matrix shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI. Among them, scar counts show significant positive correlations with both removal ratio (ρ = `r cor_scar_removal`, p = $`r p_scar_removal`$) and SDI (ρ = `r cor_scar_sdi`, p = $`r p_scar_sdi`$). Removal ratio and SDI exhibit a stronger and more highly significant correlation (ρ = `r cor_removal_sdi`, p = $`r p_removal_sdi`$). These two variables incorporate surface area in their calculation, making them more effective in capturing the density and intensity of reduction, and therefore more reliable indicators of the shaping process in spheroids.

Among the shape descriptors of spheroids, several significant correlations were observed. Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with SPHARM power (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms. As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids [citation]. This correlation between edge angle and sphericity may help explain why spheroids were often interpreted as exhausted cores in earlier research [citation]. Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult. However, the reverse is not necessarily true—not all exhausted cores evolve into spheroids.

Curvature was only significantly correlated with sphericity (ρ = `r cor_cur_sph`, p = $`r p_cur_sph`$), suggesting that smoother surfaces tend to approximate spheres more closely. Its lack of correlation with other variables may reflect its focus on local surface texture, while edge angle and SPHARM power capture broader aspects of shape regularity and symmetry. The weak correlation between sphericity and SPHARM power may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SPHARM power responds more to asymmetry and surface irregularity. Thus, even in similarly spherical specimens, surface differences can produce substantial variation in SPHARM power. The analysis of spheroid shape variables indicates that mean edge angle, curvature, sphericity, and SPHARM power each capture distinct aspects of spheroid morphology, which explains the varying strengths of correlation observed among them.
